An event-sourced and decentralized database with IPLD and materialized views

### 2019-09-26 draft.tex
update: clarify single-writer benefits + orbitdb ref
Signed-off-by: Carson Farmer <carson.farmer@gmail.com>

\subsection{ Event Logs}

@@

In multi-writer systems, determining causal order in all replicas is particularity challenging. As previously discussed (Section \ref{sec:merkleclocks}), a multi-writer solution, such as \texttt{ipfs-log}\footnote{\url{https://github.com/orbitdb/ipfs-log}}, creates a Merkle-Clock that is only partially ordered. That is, this approach cannot achieve a total order of events without implementing a data-layer conflict resolution strategy \cite{sanjuanMerkleCRDTs2019}:

Multi-writer systems based on linked log events [MISSING] with relaxed conssitency [CAP] (which is common in p2p systems) can lead to situtaitons in which merges are imppossible due to computational or network-based restrictions. For example, any one missing event may cause a branching or split, leading to (effectively) independent event 'trees'.

Our solution to deal with divergent Merkle-Clocks is to institute a \emph{single-writer rule}: A log can only be updated by a single replica or \emph{identity}. An Event Log is a \emph{single-writer Merkle-Clock} that can be totally ordered (Figure \ref{fig:EventLog}). \emph{Separate} Event Logs can be composed into advanced structures, including CRDTs \cite{enesSingleWriterPrincipleCRDT2017}.

in multi-writer logs we solve this by having linked-data in single-writer logs

in multi-writer systems, conflicts arise along the way
in order to proceed, we need to define ways to deal with conflits
with imperfect information (which is common) this inevitably leads to situations where we need to decide how to reconsile the conflict
this might include a merge, etc

in single writer systems, conflict resolution is deferred to a point at which a decision is required
this means that imperfect information may be supplemented along the way, without causing conflucts
it also means that applications can choose conflict resolution strategies specific to the task at hand
if using crdt, conflicts are irrelevant
in some instances, an additional clock may be required, such as vector or bloom clock to deal with total ordering
finally, even merges are possible if this is the desired confluct resolution strategy

@@

With such a simple interface to such a complex underlying process, it is not difficult to imagine even higher-level APIs in which Threads are exposed via interfaces compatible with \emph{existing} datastores and/or communications systems. For example, a key-value store built on Threads would ``map'' key-value operations, such as \texttt{Put}, \texttt{Get}, and \texttt{Del} to an internal (i.e., private) Model as in the previous section, with similarly defined methods. The generated Events would then mutate the internal (map-like) view Model effectively encapsulating the entire Event Store in a database structure that satisfies a given interface (see Figure \ref{ex:KVStore} for example). These too would be distributed as Modules, making it easy for developers to swap in/substitute existing backend infrastructure.

>>>>

\subsection{Event Logs}

In multi-writer systems, conflicts arise as disparate peers end up producing conflicting state changes, or changes that end up out of sync. In order to proceed, there must be some way to deal with these conflits. As mentioned previously (see Section \ref{sec:merkleclocks}), one way to avoid conflicts to some degree is to determine the total order of events in the shared log. In multi-writer systems, determining this causal order in all replicas is particularity challenging. In some cases (e.g, \texttt{ipfs-log} \cite{markroberthendersonOrbitDBFieldManual2019}), a Merkle-Clock can be used to induce ordering. However, this is only \emph{partially ordered}. That is, this approach cannot achieve a total order of events without implementing a data-layer conflict resolution strategy \cite{sanjuanMerkleCRDTs2019}:

@@

Clearly, multi-writer systems with relaxed conssitency \cite{brewerRobustDistributedSystems2000,gilbertBrewerConjectureFeasibility2002a} and imperfect information (e.g., caused by computational or network issues, both of which are common in p2p systems) will almost inevitably lead to situations where we need to decide how to reconsile a conflict. Some systems come with conflict resolution strategies ``built in'' (e.g., \texttt{ipfs-log} uses a built-in CRDT). However, this is not ideal, because it forces developers/users into a very specific usage pattern. It is particularly limiting in situations where developers might like to institute their own CRDT or domain-specific conflict resolution strategies.

Our solution to dealing with log conflicts (i.e., divergent Merkle-Clocks) is to institute a \emph{single-writer rule}: A Log can only be updated by a single replica or \emph{identity}. An Event Log is then a \emph{single-writer Merkle-Clock} that can be totally ordered (Figure \ref{fig:EventLog}). \emph{Separate} Event Logs can be composed into advanced structures, including CRDTs \cite{enesSingleWriterPrincipleCRDT2017}.

This presents a novel take on multi-writer systems \cite{enesSingleWriterPrincipleCRDT2017}, such that conflict resolution is deferred to a point at which a decision is actually \emph{required}. This means that imperfect information may be supplemented along the way (i.e., at a later time), without causing conflicts in the mean time. It also means that applications can choose conflict resolution strategies specific to the task at hand. For example, if using a downstream CRDT, ordering is actually irrelevant, and can be ignored completely. Alternatively, an additional clock may be required to ensure consistent ordering, such as vector or bloom clock (see Section \ref{sec:LogicalClocks}). Finally, even manual merge-type strategies are possible if this is the desired conflict resolution strategy.

@@

With such a simple interface to such a complex underlying process, it is not difficult to imagine even higher-level APIs in which Threads are exposed via interfaces compatible with \emph{existing} datastores and/or communications systems. Here we draw inspiration from similar projects (e.g., OrbitDB \cite{markroberthendersonOrbitDBFieldManual2019}) which have made it much easier for developers familiar with centralized database systems to make the move to decentralized systems such as Threads. For example, a key-value store built on Threads would ``map'' key-value operations, such as \texttt{Put}, \texttt{Get}, and \texttt{Del} to an internal (i.e., private) Model as in the previous section, with similarly defined methods. The generated Events would then mutate the internal (map-like) view Model effectively encapsulating the entire Event Store in a database structure that satisfies a given interface (see Figure \ref{ex:KVStore} for example). These too would be distributed as Modules, making it easy for developers to swap in/substitute existing backend infrastructure.
